---
title: "DEP Case 2 — R Extraction Code"
subtitle: "Basketball Reference NBA Player Statistics (February 2026)"
author: "Immanuel Williams"
date: "2026-02-26"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Steps"
    number-sections: true
    code-fold: false
    code-tools: true
    highlight-style: github
    theme: cosmo
    self-contained: true
    df-print: kable
execute:
  eval: false
  echo: true
  warning: true
  message: true
---

# Overview {.unnumbered}

This document contains the fully annotated R extraction code for **DEP
Case 2: Basketball Reference NBA Player Statistics**. Each section
corresponds to a step in the Data Extraction Plan (see
`dep_appendix_nocode.tex`, Section B). The code is written to be run
top-to-bottom as a single workflow.

::: callout-warning
## DEP Verdict for this case
**PROCEED WITH CONDITIONS.** The Terms of Use contain a broad prohibition
on automated access, but the qualifier "in a manner that adversely impacts
site performance or access" limits its scope. The `robots.txt` actively
permits crawling of the exact pages needed and specifies a `Crawl-delay: 3`.
This decision is documented, visible, and auditable. See DEP Section B.2.
:::

---

# Step 1 — Load Packages and Set DEP Parameters

**DEP Reference:** B.1 Purpose & Source, B.3 Access & Technical Plan

All extraction parameters defined in the DEP are declared as constants
at the top of the script. This makes it immediately clear what the
extraction covers without reading through the logic.

```{r}
#| label: setup

library(rvest)    # HTML scraping
library(dplyr)    # Data manipulation
library(purrr)    # Functional iteration: map_dfr()
library(xml2)     # Saving raw HTML snapshots

# --- DEP B.3: Extraction parameters ---
# Seasons: 2000-01 through 2024-25 (the year listed = end of season)
SEASONS   <- 2001:2025

# Rate limit: 5 seconds between all requests
# DEP B.2 justification: exceeds robots.txt Crawl-delay of 3 seconds;
# keeps total traffic to 52 requests over ~4.5 minutes.
DELAY_SEC <- 5

# --- DEP B.3: Storage paths ---
RAW_DIR <- "data/raw/bball_ref/"
dir.create(paste0(RAW_DIR, "html"), recursive = TRUE, showWarnings = FALSE)
dir.create(paste0(RAW_DIR, "csv"),  recursive = TRUE, showWarnings = FALSE)

cat("Seasons to extract:", length(SEASONS), "\n")
cat("Table types: per_game, advanced\n")
cat("Total requests planned:", length(SEASONS) * 2, "\n")
cat("Estimated time at", DELAY_SEC, "sec delay:",
    round(length(SEASONS) * 2 * DELAY_SEC / 60, 1), "minutes\n")
```

---

# Step 2 — Define the Scraping Function

**DEP Reference:** B.3 Access & Technical Plan

The single-season scraper encapsulates all the parsing decisions documented
in the DEP's "Known Parsing Challenges" section. Each challenge is handled
in a labelled block so the connection to the DEP is traceable.

The function uses `tryCatch()` to log failures without aborting the full
extraction. This is important: if one season page fails, all others should
still complete.

```{r}
#| label: scrape-function

scrape_season <- function(year, type = "per_game") {

  # --- Build URL (DEP B.3: Target URLs) ---
  url <- sprintf(
    "https://www.basketball-reference.com/leagues/NBA_%d_%s.html",
    year, type
  )

  # --- DEP B.3: Rate limiting ---
  # Sys.sleep() called BEFORE the request so the delay is always honored,
  # even if the previous request errored out.
  Sys.sleep(DELAY_SEC)

  tryCatch({
    # Fetch and snapshot raw HTML (reproducibility anchor)
    page <- read_html(url)
    write_xml(page,
              sprintf("%shtml/%s_%d.html", RAW_DIR, type, year))

    # --- DEP B.3 Known Parsing Challenge: table ID ---
    # Basketball Reference uses HTML id attributes to identify stat tables.
    table_id <- if (type == "per_game") "per_game_stats" else "advanced"

    raw_table <- page |>
      html_element(paste0("#", table_id)) |>
      html_table()

    # --- DEP B.3 Known Parsing Challenge: repeated header rows ---
    # Basketball Reference injects the header row every ~20 rows for
    # readability. These rows have "Rk" in the Rk column and must be removed.
    df <- raw_table |>
      filter(Rk != "Rk")



    # Tag with season year for merging later
    df$season <- year

    cat(sprintf("  [OK] %s %d — %d player rows\n", type, year, nrow(df)))
    return(df)

  }, error = function(e) {
    # Log the failure but return NULL so map_dfr continues
    cat(sprintf("  [FAIL] %s %d — %s\n", type, year, e$message))
    return(NULL)
  })
}
```

---

# Step 3 — Extract All Seasons

**DEP Reference:** B.3 Access & Technical Plan, B.4 Scope & Sampling

`map_dfr()` applies the scraper to every season and row-binds the results.
Progress is printed for each request so long-running extractions are
easy to monitor.

::: callout-note
## Expected runtime
At 5 seconds per request × 52 requests ≈ **4.5 minutes**. Do not interrupt
the extraction mid-run; partial results would require re-running from the
beginning (or implementing a checkpoint/resume system not shown here).
:::

```{r}
#| label: extract

cat("=== Starting per-game extraction ===\n")
per_game_data <- map_dfr(SEASONS, scrape_season, type = "per_game")

cat("\n=== Starting advanced stats extraction ===\n")
advanced_data <- map_dfr(SEASONS, scrape_season, type = "advanced")

cat("\n=== Extraction complete ===\n")
cat("Per-game rows:", nrow(per_game_data), "\n")
cat("Advanced rows:", nrow(advanced_data), "\n")
```

---

# Step 4 — Validate Extraction Results

**DEP Reference:** B.5 Validation & Contingency — *"Validation checkpoints
pre-specified before extraction begins."*

All three validation checks were defined in the DEP before any data was
seen. The spot-check value (Curry 2015–16 3PA = 11.2) was identified from
publicly known statistics before running the scraper.

```{r}
#| label: validate

# --- DEP B.5: Validation Check 1 — Row counts per season ---
# Expected: 400–600 rows per season. Fewer suggests a parsing failure
# (e.g., wrong table ID, repeated-header rows not removed, etc.)
season_counts <- per_game_data |> count(season)

under_threshold <- season_counts |> filter(n < 400)
if (nrow(under_threshold) > 0) {
  stop(
    "VALIDATION FAILED: the following seasons have fewer than 400 rows:\n",
    paste(under_threshold$season, collapse = ", "),
    "\nInvestigate before proceeding to cleaning."
  )
}
cat("Check 1 passed: all seasons have >= 400 rows.\n")
cat("Season row range:", min(season_counts$n), "–", max(season_counts$n), "\n")

# --- DEP B.5: Validation Check 2 — Spot-check value ---
# Stephen Curry 2015-16: 3PA per game = 11.2 (pre-specified in DEP)
curry_val <- per_game_data |>
  filter(Player == "Stephen Curry", season == 2016) |>
  pull("3PA") |>
  as.numeric()

stopifnot(
  "Curry 2015-16 3PA mismatch — extraction or parsing error" =
    length(curry_val) == 1 && abs(curry_val - 11.2) < 0.5
)
cat("Check 2 passed: Curry 2015-16 3PA =", curry_val, "(expected ~11.2).\n")

# --- DEP B.5: Validation Check 3 — Column counts ---
pg_cols  <- ncol(per_game_data)
adv_cols <- ncol(advanced_data)
cat("Check 3: per-game columns =", pg_cols,
    "| advanced columns =", adv_cols, "\n")
if (pg_cols < 25 || adv_cols < 20) {
  warning("Fewer columns than expected — verify table parsing.")
}
cat("All validation checks complete.\n")
```

---

# Step 5 — Apply Scope Decisions

**DEP Reference:** B.4 Scope & Sampling

Pre-specified exclusion criteria are applied here as explicit, labelled
filtering steps. The criteria were committed to in the DEP before seeing
the data distribution.

```{r}
#| label: scope

# --- DEP B.4 Exclusion Criterion: minimum games threshold ---
# Players with < 20 games played are RETAINED in raw data but
# flagged so they can be excluded from regression later.
# This decision was made in the DEP — not after observing the distribution.

per_game_data <- per_game_data |>
  mutate(
    G = as.integer(G),
    exclude_regression = G < 20,  # TRUE = exclude from regression
    exclusion_reason   = ifelse(G < 20, "fewer_than_20_games", NA_character_)
  )

excluded_count <- sum(per_game_data$exclude_regression, na.rm = TRUE)
cat("Players flagged for regression exclusion (< 20 G):", excluded_count, "\n")
cat("Players retained for regression:",
    nrow(per_game_data) - excluded_count, "\n")
```

---

# Step 6 — Save Raw Extracted Data

**DEP Reference:** B.3 Access & Technical Plan — Storage plan

```{r}
#| label: save

timestamp <- format(Sys.Date(), "%Y-%m-%d")

per_game_path <- paste0(RAW_DIR, "csv/per_game_all_", timestamp, ".csv")
advanced_path <- paste0(RAW_DIR, "csv/advanced_all_", timestamp, ".csv")

write.csv(per_game_data, per_game_path, row.names = FALSE)
write.csv(advanced_data, advanced_path, row.names = FALSE)

cat("Saved per-game data to:", per_game_path, "\n")
cat("Saved advanced data to:", advanced_path, "\n")
cat("\nExtraction complete:", timestamp, "\n")
cat("Raw HTML files in:", paste0(RAW_DIR, "html/"), "\n")
```

---

# Contingency Notes {.unnumbered}

**DEP Reference:** B.5 Validation & Contingency

If Basketball Reference blocks your IP address (`HTTP 403` or no
response):

1. Wait 24 hours without retrying
2. On retry, increase `DELAY_SEC` to 10 or higher
3. If blocking persists or Sports Reference makes contact, stop
   scraping and switch to one of the documented fallbacks:
   - **Manual CSV exports:** use the site's built-in download buttons
     (permitted by ToS; no automation involved)
   - **hoopR:** `install.packages("hoopR")` — wraps the NBA Stats API;
     note this source is less stable and column names differ
   - **Kaggle snapshots:** search for Basketball Reference datasets;
     note they may not cover the full 2000–2025 window

All fallbacks have documented limitations — the DEP records these so
the limitations are visible before a crisis forces a hasty decision.

---

# Session Info {.unnumbered}

```{r}
#| label: session-info
sessionInfo()
```
