---
title: "DEP Case 1 — R Extraction Code"
subtitle: "Wikipedia U.S. Population Tables (February 2026)"
author: "Immanuel Williams"
date: "2026-02-26"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Steps"
    number-sections: true
    code-fold: false
    code-tools: true
    highlight-style: github
    theme: cosmo
    self-contained: true
    df-print: kable
execute:
  eval: false
  echo: true
  warning: true
  message: true
editor: 
  markdown: 
    wrap: 72
---

# Overview {.unnumbered}

This document contains the fully annotated R extraction code for **DEP
Case 1: Wikipedia U.S. Population Tables**. Each section corresponds to
a step in the Data Extraction Plan (see `dep_appendix_nocode.tex`,
Section A). The code is written to be run top-to-bottom as a single
workflow. Steps are numbered to match the DEP.

::: callout-important
## DEP Verdict for this case

**PROCEED.** Legal landscape is clear and unambiguous (CC BY-SA 4.0
license, public domain data, no PII, no ToS conflict). See DEP Section
A.2.
:::

------------------------------------------------------------------------

# Step 1 — Load Packages and Define the Source URL

**DEP Reference:** A.1 Purpose & Source, A.3 Access & Technical Plan

The URL is defined as a named constant at the top of the script. This
makes it easy to update if the Wikipedia article is moved, and makes the
source auditable without reading through the code.

```{r}
#| label: setup

library(rvest)    # HTML scraping: read_html(), html_table(), html_element()
library(dplyr)    # Data manipulation: filter(), mutate(), count()
library(stringr)  # String Manipulation aspects
library(xml2)     # Saving raw HTML snapshot: write_xml()

# --- DEP A.3: Source URL (the single resource being extracted) ---
# Defined once here; never hardcoded elsewhere in the script.
WIKI_URL <- paste0(
  "https://en.wikipedia.org/wiki/",
  "List_of_U.S._states_and_territories_by_historical_population"
)

# --- DEP A.3: Storage paths ---
# Each extraction run is stored in a date-stamped folder. If reviewers
# rerun the extraction on the same date, this replaces that date's folder
# so one source/date combination has one clean set of files.
RUN_DATE <- format(Sys.Date(), "%Y-%m-%d")
RUN_DIR <- file.path("data", "raw", "wikipedia", RUN_DATE)
REPLACE_EXISTING_DATA <- TRUE

if (dir.exists(RUN_DIR) && REPLACE_EXISTING_DATA) {
  unlink(RUN_DIR, recursive = TRUE)
}
dir.create(RUN_DIR, recursive = TRUE, showWarnings = FALSE)
cat("Run directory:", RUN_DIR, "\n")
```

------------------------------------------------------------------------

# Step 2 — Fetch the Page and Save the Reproducibility Anchor

**DEP Reference:** A.3 Access & Technical Plan — *"A copy of the raw
HTML page will be saved alongside the extracted CSV."*

Wikipedia is a living document. Editors can change values, add columns,
or restructure the article at any time. Saving the raw HTML at the
moment of extraction creates a **reproducibility anchor**: anyone who
later wants to verify the extracted data can compare it against the
exact HTML source that was used.

```{r}
#| label: fetch-and-snapshot

# Fetch the page (single HTTP GET request — no auth, no rate limit needed)
raw_html <- read_html(WIKI_URL)

# Save raw HTML snapshot immediately, before any parsing
# This is the reproducibility anchor described in DEP A.3.
snapshot_path <- file.path(RUN_DIR, "wikipedia_population_raw.html")
write_xml(raw_html, snapshot_path)
cat("HTML snapshot saved to:", snapshot_path, "\n")
```

------------------------------------------------------------------------

# Step 3 — Extract the Target Table

**DEP Reference:** A.3 Access & Technical Plan — *"Target table
identified by index position (third table as of January 2025). Known
fragility: if the article is reorganized, the index may silently capture
the wrong table."*

The DEP documents this fragility explicitly. The mitigation is mandatory
visual inspection in Step 4 before any downstream processing.

```{r}
#| label: extract-table

# Parse all tables on the page
all_tables <- raw_html |> html_table(fill = TRUE)
cat("Total tables found on page:", length(all_tables), "\n")

# Extract the target table by index
# DEP NOTE: This index was verified as of January 2025.
#           If the article is restructured, update this index and
#           re-verify with the spot-checks in Step 5.
TARGET_INDEX <- 4
pop_table <- all_tables[[TARGET_INDEX]]

cat("Extracted table dimensions:",
    nrow(pop_table), "rows x", ncol(pop_table), "cols\n")
```

------------------------------------------------------------------------

# Step 4 — Mandatory Visual Inspection

**DEP Reference:** A.3 Access & Technical Plan — *"The extraction script
must print the first 3 rows before downstream processing."*

This step cannot be skipped or suppressed. Its purpose is to verify that
the correct table was captured before any transformations make it harder
to detect the wrong-table error.

```{r}
#| label: inspect

# Print the first 3 rows — confirm this looks like a state-population table
# Expected: row names are state names, column names are census years
print(head(pop_table, 3))

# Also check the column names to confirm census years are present
cat("\nColumn names (first 10):\n")
print(names(pop_table)[1:10])
```

------------------------------------------------------------------------

# Step 5 — Spot-Check Validation

**DEP Reference:** A.5 Validation & Contingency — *"Pre-specified
spot-check values (from Census Bureau publications)."*

These three values were identified in the DEP **before** extraction, not
after looking at the data. This is the extraction equivalent of
preregistration: defining the validation criteria in advance removes the
possibility of post-hoc rationalization.

::: callout-note
## Why pre-specification matters

A validation checkpoint defined *after* looking at the data is
indistinguishable from confirming what you already know. Pre-specified
checkpoints constrain what counts as a passing extraction.
:::

```{r}
#| label: validate

# --- Pre-specified spot-check values (DEP A.5) ---
# Sources: 2020 Decennial Census (CA, WY)
spot_checks <- tibble(
  state    = c("California", "Wyoming"),
  expected = c(39538223L, 576851L)
)

# Value spot-checks
# These will throw an informative error if the extracted values don't match.
# Adjust the column references after cleaning column names in your workflow.
pop_table |>
  filter(.data[[names(pop_table)[1]]] %in% spot_checks$state) |>
  select(state = 1, value = `2020`) |>
  mutate(value = as.integer(str_remove_all(value, ","))) |>  # strip commas before coercion
  left_join(spot_checks, by = "state") |>
  mutate(match = value == expected) |>
  {
    \(df) {
      if (!all(df$match)) {
        mismatches <- df |> 
          filter(!match)
        stop("Spot-check failed for: ",
             paste(mismatches$state, collapse = ", "),
             " — check table index or source change")
      }
      cat("All spot-checks passed.\n")
      df
    }
  }()
```

------------------------------------------------------------------------

# Step 6 — Apply Scope Decisions and Save

**DEP Reference:** A.4 Scope & Sampling

The scope decisions documented in the DEP are applied here as explicit
filtering steps. Each filter is labelled with the DEP reference so that
the connection between plan and code is traceable.

```{r}
#| label: scope-and-save

# --- DEP A.4: Flag territories (not removed, just marked) ---
# The 50 states are included in primary analysis;
# territories are retained in raw data but flagged.
us_states <- c(
  "Alabama","Alaska","Arizona","Arkansas","California","Colorado",
  "Connecticut","Delaware","Florida","Georgia","Hawaii","Idaho",
  "Illinois","Indiana","Iowa","Kansas","Kentucky","Louisiana","Maine",
  "Maryland","Massachusetts","Michigan","Minnesota","Mississippi",
  "Missouri","Montana","Nebraska","Nevada","New Hampshire",
  "New Jersey","New Mexico","New York","North Carolina","North Dakota",
  "Ohio","Oklahoma","Oregon","Pennsylvania","Rhode Island",
  "South Carolina","South Dakota","Tennessee","Texas","Utah",
  "Vermont","Virginia","Washington","West Virginia","Wisconsin","Wyoming"
)

pop_table <- pop_table |>
  mutate(is_state = pop_table[[1]] %in% us_states)  # TRUE = state

# Confirm all 50 states present
state_count <- sum(pop_table$is_state)
if (state_count < 50) {
  warning("Fewer than 50 states found (", state_count, "). Investigate.")
}
cat("States identified:", state_count, "\n")

# --- DEP A.4: Log territory count for audit trail ---
territory_count <- nrow(pop_table) - state_count
cat("Territories/other rows flagged:", territory_count, "\n")

# --- Save extracted data ---
output_path <- file.path(RUN_DIR, "wikipedia_population.csv")
write.csv(pop_table, output_path, row.names = FALSE)
cat("\nExtraction complete:", RUN_DATE, "\n")
cat("Output saved to:", output_path, "\n")
cat("Reproducibility anchor:", snapshot_path, "\n")
```

------------------------------------------------------------------------

# Contingency Notes {.unnumbered}

**DEP Reference:** A.5 Validation & Contingency

If the spot-checks in Step 5 fail, do **not** proceed to data cleaning.
Instead:

1.  Open the saved HTML snapshot and visually confirm the table
    structure
2.  Check whether Wikipedia has updated the article (compare with a
    recent Wayback Machine snapshot at `web.archive.org`)
3.  If the article structure has changed, update `TARGET_INDEX` and
    re-run from Step 3
4.  If data values have changed (legitimate Wikipedia edits), note the
    discrepancy in your extraction log and update the spot-check values
    with the new Census Bureau-verified figures
5.  If Wikipedia is unavailable entirely, use the Census Bureau's
    historical tables at
    `census.gov/data/tables/time-series/demo/popest/`

------------------------------------------------------------------------

# Session Info {.unnumbered}

```{r}
#| label: session-info
sessionInfo()
```
