---
title: "DEP Case 3 — R Extraction Code"
subtitle: "Spotify Audio Features via Web API (September 2021)"
author: "Immanuel Williams"
date: "2021-09-01"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Steps"
    number-sections: true
    code-fold: false
    code-tools: true
    highlight-style: github
    theme: cosmo
    self-contained: true
    df-print: kable
execute:
  eval: false
  echo: true
  warning: true
  message: true
editor: 
  markdown: 
    wrap: 72
---

# Overview {.unnumbered}

This document contains the fully annotated R extraction code for **DEP
Case 3: Spotify Audio Features via Web API**. Each section corresponds
to a step in the Data Extraction Plan (see `dep_appendix_nocode.tex`,
Section C). The code is written to be run top-to-bottom as a single
workflow.

::: callout-warning
## Temporal context — read before running

This DEP and code were written as if it were **September 2021**. Since
then:

-   **November 2024:** Spotify deprecated the Audio Features and Audio
    Analysis endpoints for new developer applications.
-   **March 2025:** Extended API access restricted to apps with ≥250,000
    monthly active users and a registered business.
-   **February 2026:** Developer Mode requires a Premium subscription;
    limited to 5 users.
-   **Late 2024:** `spotifyr` was removed from CRAN following the API
    deprecation.

**This code will not run with a new Spotify developer account as of
2025.** It is preserved here to document the extraction process and
validate that all four contingency scenarios in DEP C.5 materialized. To
work with Spotify-like audio features today, use the fallbacks
documented in DEP C.5: MusicBrainz, ListenBrainz, or Last.fm.
:::

::: callout-note
## DEP Verdict for this case (September 2021)

**PROCEED** with documented ambiguity about the "train machine learning
models" clause. The researcher interprets academic multiple linear
regression as distinct from building a competing recommendation service.
This interpretation is recorded, not assumed. See DEP Section C.2.
:::

------------------------------------------------------------------------

# Step 1 — Load Packages and Authenticate

**DEP Reference:** C.3 Access & Technical Plan — *"Authentication: OAuth
2.0. Client ID and Client Secret stored as environment variables. Never
hardcoded in scripts."*

Credentials are read from environment variables, not from a
configuration file checked into version control. This is non-negotiable:
hardcoded API keys in scripts are a security vulnerability and a
reproducibility hazard.

::: callout-tip
## Setting credentials safely

Before running this script, set your credentials in your `.Renviron`
file (never in the script itself):

```         
SPOTIFY_CLIENT_ID=your_client_id_here
SPOTIFY_CLIENT_SECRET=your_client_secret_here
```

Then restart R. Use `usethis::edit_r_environ()` to open the file.
:::

```{r}
#| label: setup

# DEP C.5 Contingency 3: If spotifyr is removed from CRAN,
# pin the current version using renv.
# Run renv::snapshot() after installing to lock this version.
library(spotifyr)   # Thompson et al., v2.2.4 — see DEP C.1
library(dplyr)
library(purrr)

# --- DEP C.3: Authenticate using environment variables only ---
# If credentials are missing, stop with an informative message.
if (Sys.getenv("SPOTIFY_CLIENT_ID") == "") {
  stop(
    "SPOTIFY_CLIENT_ID not found in environment.\n",
    "Set it in .Renviron and restart R. See DEP C.3 for details."
  )
}

access_token <- get_spotify_access_token()
cat("Authentication successful.\n")

# --- DEP C.3: Storage paths ---
# Each extraction run is stored in a date-stamped folder. If reviewers
# rerun the extraction on the same date, this replaces that date's folder
# so one source/date combination has one clean set of files.
RUN_DATE <- format(Sys.Date(), "%Y-%m-%d")
RUN_DIR <- file.path("data", "raw", "spotify", RUN_DATE)
REPLACE_EXISTING_DATA <- TRUE

if (dir.exists(RUN_DIR) && REPLACE_EXISTING_DATA) {
  unlink(RUN_DIR, recursive = TRUE)
}
dir.create(RUN_DIR, recursive = TRUE, showWarnings = FALSE)
cat("Run directory:", RUN_DIR, "\n")
```

------------------------------------------------------------------------

# Step 2 — Define the Artist List (Sampling Strategy)

**DEP Reference:** C.4 Scope & Sampling — *"Stratified sampling:
complete discographies for the top 10 artists by Spotify monthly
listeners in each of 6 genres."*

Artist IDs are resolved and recorded in the DEP before extraction
begins, not looked up dynamically during the run. This makes the sample
definition auditable and stable.

::: callout-important
## Documented sampling bias (DEP C.4)

Spotify's Browse and Search endpoints return results ranked by
popularity. There is no mechanism for random artist retrieval via the
API. The sample is therefore biased toward the high end of the
popularity distribution.

This is a property of the data **source**, not solely a choice of the
analyst. The DEP records this distinction explicitly so that readers of
the eventual analysis understand the provenance of the range
restriction.
:::

```{r}
#| label: artist-list

# Spotify Artist IDs for top-10 artists per genre
# (resolved before extraction — not dynamically queried)
# Full list; shown here for pop and rock as illustration.

artist_ids <- list(
  pop = c(
    "06HL4z0CvFAxyc27GXpf02",   # Taylor Swift
    "1uNFoZAHBGtllmzznpCI3s",   # Justin Bieber
    "3TVXtAsR1Inumwj472S9r4",   # Drake
    "4dpARuHxo51G3z768sgnrY",   # Adele
    "6eUKZXaKkcviH0Ku9w2n3V",   # Ed Sheeran
    "7dGJo4pcD2V6oG8kP0tJRR",   # Eminem
    "1Xyo4u8uXC1ZmMpatF05PJ",   # The Weeknd
    "66CXWjxzNUsdJxJ2JdwvnR",   # Ariana Grande
    "0du5cEVh5yTK9QJze8zA0C",   # Bruno Mars
    "3WrFJ7ztbogyGnTHbHJFl2"    # The Beatles
  ),
  rock = c(
    "22bE4uQ6baNwSHPVcDxLCe",   # The Rolling Stones
    "36QJpDe2go2KgaRleHCDTp",   # Led Zeppelin
    "711MCceyCBcFnzjGY4Q7Un",   # AC/DC
    "5K4W6rqBFWDnAN6FQUkS6x",   # Kanye West (crossover)
    "12Chz98pHFMPJEknJQMWqN",   # Metallica
    "0k17h0D3J5VfsdmQ1iZtE9",   # Pink Floyd
    "0oSGxfWSnnOXhD2fKuz2Gy",   # David Bowie
    "1Cs0zKBU1kc0i8ypK3B9ai",   # Radiohead
    "7u6zL7kqpgLPiNbuCxsdAL",   # Arctic Monkeys
    "3YQKmKGau1PzlVlkL1iAx7"    # Tame Impala
  )
  # hip-hop, country, electronic, r-and-b: full list in supplementary
  # materials. Pattern is identical — replace artist IDs only.
)

total_artists <- sum(lengths(artist_ids))
cat("Total artists to extract:", total_artists, "\n")
cat("Genres:", paste(names(artist_ids), collapse = ", "), "\n")
```

------------------------------------------------------------------------

# Step 3 — Define the Extraction Function

**DEP Reference:** C.3 Access & Technical Plan — rate limiting, error
handling, exclusion criteria

```{r}
#| label: extract-function

get_artist_features <- function(artist_id, genre) {

  # --- DEP C.3: Rate limiting ---
  # Empirically determined limit ~100 req/min (see DEP C.3).
  # 0.5 sec delay keeps us well under that threshold.
  Sys.sleep(0.5)

  tryCatch({
    feats <- get_artist_audio_features(
      artist         = artist_id,
      authorization  = access_token,
      include_groups = "album"  # DEP C.4 Exclusion: compilations excluded
                                # by requesting only "album" type
    )

    # Tag with genre for stratified analysis
    feats$genre <- genre
    cat(sprintf("  [OK] %s (%s) — %d tracks\n",
                feats$artist_name[1], genre, nrow(feats)))
    return(feats)

  }, error = function(e) {
    # HTTP 429 (Too Many Requests) or other error: log and continue
    cat(sprintf("  [FAIL] %s (%s) — %s\n", artist_id, genre, e$message))
    return(NULL)
  })
}
```

------------------------------------------------------------------------

# Step 4 — Run the Extraction

**DEP Reference:** C.4 Scope & Sampling

`map2_dfr()` iterates over the artist IDs and genre labels in parallel,
calling the extraction function for each artist and binding results
row-wise into a single data frame.

```{r}
#| label: extract

# Build parallel vectors of (artist_id, genre) for map2_dfr
genre_labels <- rep(names(artist_ids), lengths(artist_ids))
artist_vec   <- unlist(artist_ids)

cat("=== Starting Spotify extraction ===\n")
cat("Total artists:", length(artist_vec), "\n\n")

all_features <- map2_dfr(genre_labels, artist_vec,
                         ~ get_artist_features(.y, .x))

cat("\n=== Extraction complete ===\n")
cat("Total tracks extracted:", nrow(all_features), "\n")
```

------------------------------------------------------------------------

# Step 5 — Apply Exclusion Criteria

**DEP Reference:** C.4 Scope & Sampling — *"Exclusion criteria
pre-specified: compilation albums; tracks with popularity = 0; tracks
missing audio features."*

Each exclusion is applied sequentially with a count logged before and
after. This produces the audit trail required by the DEP.

```{r}
#| label: exclusions

n_start <- nrow(all_features)
cat("Tracks before exclusions:", n_start, "\n\n")

# --- DEP C.4 Exclusion 1: Popularity = 0 ---
# Indicates inactive or removed content.
excl_zero_pop <- all_features |> filter(track_popularity == 0)
cat("Excluded (popularity = 0):", nrow(excl_zero_pop), "\n")
all_features <- all_features |> filter(track_popularity > 0)

# --- DEP C.4 Exclusion 2: Missing audio features ---
# Any track missing values in the 13 feature columns is excluded.
audio_feature_cols <- c(
  "danceability", "energy", "loudness", "speechiness",
  "acousticness", "instrumentalness", "liveness", "valence",
  "tempo", "duration_ms", "time_signature", "key", "mode"
)
excl_missing <- all_features |>
  filter(if_any(all_of(audio_feature_cols), is.na))
cat("Excluded (missing audio features):", nrow(excl_missing), "\n")
all_features <- all_features |>
  filter(!if_any(all_of(audio_feature_cols), is.na))

n_end <- nrow(all_features)
cat("\nTracks after all exclusions:", n_end, "\n")
cat("Total excluded:", n_start - n_end, "\n")
```

------------------------------------------------------------------------

# Step 6 — Validate Extraction Results

**DEP Reference:** C.5 Validation & Contingency — *"Validation
checkpoints pre-specified before extraction begins."*

```{r}
#| label: validate

cat("=== Running validation checks ===\n\n")

# --- DEP C.5: Check 1 — Minimum tracks per artist ---
artist_counts <- all_features |> count(artist_name)
under_minimum <- artist_counts |> filter(n < 20)
if (nrow(under_minimum) > 0) {
  warning(
    "The following artists have fewer than 20 tracks — ",
    "likely a retrieval error:\n",
    paste(under_minimum$artist_name, collapse = ", ")
  )
} else {
  cat("Check 1 passed: all artists have >= 20 tracks.\n")
}

# --- DEP C.5: Check 2 — Range checks on bounded audio features ---
bounded_cols <- c("danceability", "energy", "speechiness",
                  "acousticness", "instrumentalness", "liveness", "valence")
for (col in bounded_cols) {
  out_of_range <- sum(
    all_features[[col]] < 0 | all_features[[col]] > 1,
    na.rm = TRUE
  )
  if (out_of_range > 0) {
    stop("VALIDATION FAILED: ", out_of_range,
         " out-of-range values in column: ", col)
  }
}
cat("Check 2 passed: all bounded audio features in [0, 1].\n")

# --- DEP C.5: Check 3 — Tempo range ---
tempo_violations <- sum(
  all_features$tempo < 40 | all_features$tempo > 220,
  na.rm = TRUE
)
if (tempo_violations > 0) {
  warning(tempo_violations, " tracks with tempo outside 40–220 BPM.")
} else {
  cat("Check 3 passed: all tempo values in 40–220 BPM.\n")
}

# --- DEP C.5: Check 4 — Popularity is integer in [0, 100] ---
stopifnot(
  "Popularity out of [0, 100]" =
    all(all_features$track_popularity >= 0 &
        all_features$track_popularity <= 100,
        na.rm = TRUE)
)
cat("Check 4 passed: all popularity values in [0, 100].\n")

# --- DEP C.5: Check 5 — Duplicate track IDs ---
n_dupes <- sum(duplicated(all_features$track_id))
if (n_dupes > 0) {
  warning(n_dupes, " duplicate track IDs detected — investigate ",
          "before analysis (may indicate artists shared across genres).")
} else {
  cat("Check 5 passed: no duplicate track IDs.\n")
}

cat("\nAll validation checks complete.\n")
```

------------------------------------------------------------------------

# Step 7 — Save in Two Formats

**DEP Reference:** C.3 Access & Technical Plan — *"Two formats per
extraction: `.rds` (R native, preserving data types) for analysis;
`.csv` for portability."*

```{r}
#| label: save

rds_path <- file.path(RUN_DIR, "spotify_all.rds")
csv_path <- file.path(RUN_DIR, "spotify_all.csv")

saveRDS(all_features, rds_path)
write.csv(all_features, csv_path, row.names = FALSE)

cat("Saved .rds to:", rds_path, "\n")
cat("Saved .csv to:", csv_path, "\n")
cat("\nExtraction summary:\n")
cat("  Tracks:     ", nrow(all_features), "\n")
cat("  Variables:  ", ncol(all_features), "\n")
cat("  Genres:     ", length(unique(all_features$genre)), "\n")
cat("  Artists:    ", length(unique(all_features$artist_name)), "\n")
cat("  Date:       ", RUN_DATE, "\n")
```

------------------------------------------------------------------------

# Contingency Notes {.unnumbered}

**DEP Reference:** C.5 Validation & Contingency

As documented in the DEP, four contingency scenarios were identified in
September 2021. All four materialized between 2024 and 2026:

**Scenario 1 — Rate limits tighten:** Increase `Sys.sleep()` from 0.5 to
1.0 or 2.0 seconds. Spread extraction across multiple sessions using a
checkpoint file that records which artist IDs have already been
processed.

**Scenario 2 — Audio Features endpoint deprecated (MATERIALIZED Nov.
2024):** Fall back to the Kaggle "Spotify Tracks Dataset" snapshot.
Document this substitution in your analysis with the limitations noted
in DEP C.1 (no DEP for the Kaggle dataset; provenance unclear).

**Scenario 3 — `spotifyr` removed from CRAN (MATERIALIZED):** Restore
the pinned version using:

``` r
renv::restore()   # requires renv.lock created at time of extraction
```

If no lock file exists, install from GitHub:

``` r
remotes::install_github("charlie86/spotifyr@v2.2.4")
```

**Scenario 4 — Spotify developer account suspended (effectively
materialized):** Pivot to one of these alternative audio metadata
sources:

| Source | Programmatic Access | Coverage | Notes |
|----------------|------------------|----------------|----------------------|
| MusicBrainz | Free REST API | \~2M releases | No audio features; metadata only |
| ListenBrainz | Free REST API | Listening data | Community-submitted |
| Last.fm | Free API | \~185M tracks | Scrobble-based features |
| AcousticBrainz | Static dataset (archive) | \~2M recordings | Deprecated; archived only |

------------------------------------------------------------------------

# Session Info {.unnumbered}

```{r}
#| label: session-info
sessionInfo()
```
